---
title: "Problem Set 4"
subtitle: "Due date: 9 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu) and to your class Github repository. You need to upload both your Quarto document (with R code, as needed) and its rendered output (either as a PDF or HTML file).

**Total points: 30**

## Question 1

*Points: 5*

**Suppose I am playing poker with a regular deck of 52 cards. I have three cards in my hand, one of which is a king. What is the probability that the next two cards I draw will both be kings?**

```{r}
q1_prob_a <- 3/49
q1_prob_b <- 2/48
q1_prob_a*q1_prob_b
```

The probability that the next two cards drawn will both be kings is 0.00255102 or, in other words, there is .255% chance the next two cards drawn will both be kings.

## Question 2

*Points: 5*

**A survey of 1,500 randomly selected individuals reveals that 46% of them have a college degree and that 31% of them have a gun in their house. If you were to assume that college graduation and gun ownership are independent of one another, what is the probability that an individual is both a college graduate and does not have a gun in their house?**

```{r}
q2_prob_a <- .46
1-.31
q2_prob_b <- .69
q2_prob_a*q2_prob_b
```

The probability that an individual is both a college graduate and does not have a gun in their house is 0.3174 or, in other words, there is a 31.74% chance that an individual is both a college graduate and does not have a gun in their house.

## Question 3

*Points: 5*

**I roll a fair six-sided dice and I get a number greater than three. What is the probability that the next two dice rolls will also generate numbers greater than three?**

```{r}
q3_prob_a <- .5
q3_prob_b <- .5
q3_prob_a*q3_prob_b
```

The probability that the next two dice rolls will also generate numbers greater than three is 0.25 or 25%.

## Question 4

*Points: 10*

**Assume that in the United States, 15% of people are smokers and, of smokers, 60% smoke at least one pack a day. Among those who smoke at least a pack a day, 60% will develop lung cancer at some point in their lives while, in the rest of the population (which includes smokers who smoke less than a pack a day and non-smokers), only 10% will. What is the probability that a person who develops lung cancer was a smoker who smoked at least a pack a day?**

The probability that a person who develops lung cancer was a smoker who smoked at least a pack a day is .06 or 6%.

```{r}
q4_prob_a <- .6*.1
q4_prob_b <- .15*.6
q4_prob_a*q4_prob_b/q4_prob_b
```

## Question 5

*Points: 5*

**Assume that in the world, 60% of countries are democracies and 80% of countries have ratified the Rome Statute for the International Criminal Court. Assume further that 90% of democracies have ratified the Rome Accords. What is the probability that a country is either a democracy or has ratified the Rome Accords?**

```{r}
q5_prob_a <- .6
q5_prob_b <- .8
q5_prob_b_given_a <- q5_prob_a*q5_prob_b
q5_prob_a + q5_prob_b - q5_prob_b_given_a
```

The probability that a country is either a democracy or has ratified the Rome Accords is .92 or 92%.
